Some examples in extended logic counting classes

### 2017-01-09 LICS/beyond.tex

In the previous sections, we use weighted logics to give a framework for descriptive complexity of functions. Here, we go beyond weighted logics to add unexplored quantifiers at the quantitative level. Specifically, we give the first steps on defining quantitative recursion. This goal is not trivial for two reasons: (1) we want to add recursion over functions and (2) it is not clear what could be the right notion of ``fixed point'' . 
Towards this goal, we show first how to extend $\qso$ with function symbols to later use them to define recursion at a quantitative level. 
Furthermore, we define a natural generalization of (boolean) least fixed point for functions that we called \emph{least s-fixed point} and, as a proof of concept, we show that it captures $\fp$.
Finally,we use our new tool to define an operator for counting paths in a graph, a natural generalization of the transitive closure operator~\cite{immerman1999descriptive}, and show that this captures the expressiveness of~$\shl$.

For defining recursion, we first need to define an extension of $\qso$ with function symbols. Assume that $\fs$ is an infinite set of function symbols, where each $h \in \fs$ has an associated arity, which is denoted by $\arity(h)$. Then the set of $\fqso$ formulas over a relational signature $\R$ is defined by the following grammar:
\begin{multline}
\label{eq-fqso}
	\alpha := \varphi \ \mid \  s \  \mid \  h(x_1, \ldots, x_\ell) \  \mid \
	(\alpha \add \alpha) \  \mid\  (\alpha \mult \alpha) \  \mid \\  
	\sa{x} \alpha \  \mid \
	\pa{x} \alpha \  \mid \
	\sa{X} \alpha \  \mid \
	\pa{X} \alpha,
\end{multline}
where $h \in \fs$, $\arity(h) = \ell$ and $x_1, \ldots, x_\ell$ is a sequence of (non-necessarily distinct) first-order variables. Given an $\R$-structure $\A$ with domain $A$, we say that $F$ is a \emph{function assignment} for $\A$ if for every $h \in \fs$ with $\arity(h) = \ell$, we have that $F(h) :  A^\ell \to \N$. The notion of function assignment is used to extend the semantics of $\qso$ to the case of a quantitative formula of the form $h(x_1, \ldots, x_\ell)$. More precisely, given first-order and second-order assignments $v$ and $V$ for $\A$, respectively, 
we have that:
\begin{eqnarray*}
\sem{h(x_1, \ldots, x_\ell)}(\A,v,V,F) & = & F(h)(v(x_1),\ldots, v(x_\ell)).
\end{eqnarray*}
As for the case of $\qfo$, we define $\fqfo$ disallowing quantifiers $\Sigma X$ and $\Pi X$ in \eqref{eq-fqso}.

It worths noting that function symbols represent in $\fqso$ functions from tuples to numbers and cannot be compared to the classical notion of (boolean) function symbol in first-order logics~\cite{enderton2001mathematical}. 
Furthermore, one could see a function symbol as an ``oracle'' that is instantiated by the function assignment. 
As far as we know, this is the first paper proposing this extension on weighted logics and it would be interesting to study this concept further (i.e. outside recursion).

We define an operator which extends Least Fixed Point Logic (LFP) \cite{I86,vardi1982complexity} over $\fo$ to allow counting. 
%We show how to define this operator in $\qfo(\fo)$ and we discuss how to extend these ideas to $\qso$ towards the end of the section. 
Fix a relational signature $\R$. Then the set of $\rqfo(\fo)$ formulas (Recursive $\qfo$) over $\R$ is defined as an extension of $\qfo(\fo)$ by adding the operator $\clfp{\beta(\x, h)}$ where $\x = (x_1, \ldots, x_\ell)$ is a sequence of $\ell$ distinct first-order variables and $\beta(\x, h)$ is an $\fqfo(\fo)$-formula over $\R$ whose only function symbol is $h$, with $\arity(h) = \ell$. The free variables of the formula $\clfp{\beta(\x,h)}$ are $x_1, \ldots, x_\ell$; in particular, $h$ is not considered to be free in this formula.

Fix an $\R$-structure with domain $A$ and a quantitative formula $\clfp{\beta(\x,h)}$, and assume that $\F$ is the set of functions $f :A^\ell \to \N$. To define the semantics of $\clfp{\beta(\x,h)}$, we first show how $\beta(\x,h)$ can be interpreted as an operator $T_{\beta}$ on $\F$. More precisely, for every $f \in \F$ and tuple $\a = (a_1, \ldots, a_{\ell}) \in A^\ell$, the function $T_{\beta}(f)$ satisfies that:
\begin{eqnarray*}
T_{\beta}(f)(\a) & = & \sem{\beta(\x, h)}(\A,v,F),
\end{eqnarray*}
where $v$ is a first-order assignment  for $\A$ such that $v(x_i) = a_i$ for every $i \in \{1, \ldots, \ell\}$, and $F$ is a function assignment for $\A$ such that $F(h) = f$.

@@

\begin{example} \label{ex:count-path}
%As an example, 
We would like to define an $\rqfo(\fo)$-formula that, given a directed acyclic graph $G$ with $n$ nodes and a pair of nodes $b$, $c$ in $G$, counts the number of paths of length at most $n$ from $b$ to $c$ in $G$. To this end, assume that graphs are encoded using the relational signature $\R = \{ E(\cdot,\cdot) \}$, and then define formula $\alpha(x, y, f)$ as follows:
$$
%\alpha(x,y,R,\pi) = 
%(\neg \exists zR(z))\cdot(x = y) 
E(x,y) + \sa{z} f(x,z)\cdot E(z,y).
$$
We have that $\clfp{\alpha(x,y,f)}$ defines our counting function. In fact, assume that $\A$ is an $\R$-structure with $n$ elements in its domain encoding an acyclic directed graph. Moreover, assume that $b,c$ are elements of $\A$ and $v$ is a first-order assignment over $\A$ such that $v(x) = b$ and $v(y) = c$. Then we have that $\sem{\alpha(x,y,f)}(\A,v)$ is equal to the  number of paths in $\A$ from $b$ to $c$ of length at most $n$.

Assume now that we need to extend our previous counting function to the case of arbitrary directed graphs. To this end, suppose that $\varphi_{\text{\rm first}}(x)$ and $\varphi_{\text{succ}}(x,y)$ are the $\fo$-formulas for defining the first and successor predicates, respectively, of $<$. Moreover, define formula $\beta(x, y, t, g)$ as follows:
\begin{multline*}
(E(x,y) + \sa{z} g(x,z,t)\cdot E(z,y)) \cdot \varphi_{\text{\rm first}}(t) \ +\\
\sa{t'} \varphi_{\text{succ}}(t',t) \cdot \left(\sa{x'} \sa{y'} g(x',y',t') \right)
\end{multline*}
Then our extended counting function is defined by the following formula:
$$
\sa{t} (\varphi_{\text{\rm first}}(t) \wedge \clfp{\beta(x,y,t,g)}).
$$ 
In fact, the number of paths of length at most $n$ from a node $x$ to a node $y$ is recursively computed by using the formula $(E(x,y) + \sa{z} g(x,z,t)\cdot E(z,y)) \cdot \varphi_{\text{\rm first}}(t)$, which stores this value in $g(x,y,t)$ with $t$ the first element in the domain.  The other formula, the one with the filter $\varphi_{\text{succ}}(t',t)$, is used as a counter that allows to reach a fixed point in the support of function $g$ in $n$ steps.
\end{example}
%
%
%It is well known that least fixed point logic is contained in second-order logic \cite{L04}. In the following theorem we show that the same holds in our case.
%\begin{theorem} \label{so-rec}
%$\rqfo \subseteq \qso$
%%	Given a positive $\fo$ formula $\varphi(\bar{x},R)$ and a $\qfo$ formula $\alpha(\bar{x})$, there exists a $\qso$ formula $\beta(\bar{x})$ such that $\sem{[\alfp\varphi(\bar{x},R)\mid \alpha(\bar{x},R)](\bar{x})} = \sem{\beta(\bar{x})}$.
%\end{theorem} 
Note that, in contrast to $\lfp$-logic, here we do not need to impose any positive restriction to the formula $\beta(\x,h)$.
Indeed, since $\beta$ is constructed from monotones operations (i.e. sum and product) over the natural numbers, the resulting operator $T_{\beta}$ is monotone as well.

Now that a least fixed point operator over functions is defined, the next step is to understand its expressive power.
In the following result we show how lsfp-operator can be used to capture all functions that can be computed in polynomial time.
\begin{theorem} \label{rqfo-fo-cap}
	$\rqfo(\fo)$ captures $\fp$ over the class of ordered structures.
\end{theorem}

@@

Our last goal in this section is to use the new characterization of $\fp$ to explore classes below it.
It was shown in \cite{I86,I88} that $\fo$ extended with a transitive closure operator captures $\nlog$. 
%Moreover, it was shown in \cite{I83} that $\so$ extended with a transitive closure operator captures $\pspace$. 
Inspired by this work, we show that a restricted version of $\rqfo$ can be used to capture $\shl$, the counting version of $\nlog$. 
Specifically, we use $\rqfo$ to define an operator for counting the number of paths in a directed graph which is exactly what we need for $\shl$. 
%Besides, we show that the same idea can be used to extend $\qso$ allowing to capture harder complexity classes.

Given a relation signature $\R$, the set of transitive $\qfo$ formulas ($\tqfo$-formulas) is defined as an extension of $\qfo$ with the operator
%\begin{multline}
%	\label{eq-def-tqso}
%	\alpha := \varphi \, \mid \, s \, \mid \, (\alpha \add \alpha) \, \mid\, (\alpha \mult \alpha) \, \mid \, 
%	\sa{x} \alpha \, \mid\, \
%	\pa{x} \alpha \, \mid \\ 
%	\sa{X} \alpha \, \mid \, \pa{X} \alpha \, \mid \, [\pth \psi(\bar{x}, \bar{X},\bar{y}, \bar{Y})],
%\end{multline}
$
[\pth \psi(\bar{x},\bar{y})]
$ where $\psi(\x, \y)$ is an $\fo$-formula over $\R$, $\bar{x} = (x_1, \ldots, x_k)$, $\bar{y} = (y_1, \ldots, y_k)$ are tuples of pairwise distinct first-order variables. The semantics of $[\pth \psi(\bar{x},\bar{y})]$ can easily be defined in terms of $\rqfo(\fo)$ as follows. 
Given an $\R$-structure $\A$ with domain $A$, define a (directed) graph $\cG_{\psi}(\A) = (N,E)$ such that $N = A^k$ and for every pair $\bar b, \bar c \in N$, it holds that $(\bar b, \bar c) \in E$ if, and only if, $\A \models \psi(\bar b, \bar c)$.
Similar than for Example~\ref{ex:count-path}, we can count the paths of length at most $|A^k|$ with the formula $\beta_{\psi(\bar{x},\bar{y})}(\x, \y, \t, g)$:
\begin{multline*}
(\psi(\bar{x},\bar{y}) + \sa{\z} g(\x,\z,\t)\cdot \psi(\z,\y)) \cdot \varphi_{\text{\rm first-lex}}(\t) \ +\\
\sa{\t'} \varphi_{\text{succ-lex}}(\t',\t) \cdot \left(\sa{\x'} \sa{\y'} g(\x',\y',\t') \right)
\end{multline*}
where $\varphi_{\text{\rm first-lex}}$ and $\varphi_{\text{succ-lex}}$ are $\fo$-formulas for defining the first and successor predicates over tuples in $A^k$, following the lexicographic order induced by~$<$.
Then the semantics of the path operator is defined $\rqfo$ formula:
\[
[\pth \psi(\bar{x}, \bar{y})] := \sa{\t} (\varphi_{\text{\rm first}}(\t) \wedge \clfp{\beta_{\psi(\bar{x},\bar{y})}(\x,\y,\t,g)})
\]
In other words, $\sem{[\pth \psi(\bar{x}, \bar{y})]}(\A,v,V)$ counts the number of paths in the graph $\cG_{\psi}(\A)$ whose length is at most~$|A^k|$.
%As for the case of $\qso$, the logic $\tqso(\LL)$ is obtained by restricting $\varphi$ in \eqref{eq-def-tqso} to be a formula in $\LL$. Moreover, the logic $\tqfo$ is obtained by disallowing in \eqref{eq-def-tqso} formulas $\sa{X} \alpha$ and $\pa{X} \alpha$, and by only allowing  first-order free-variables in the formula $\psi$ used in $[\pth \psi]$ in \eqref{eq-def-tqso}. 
As it was previously said, the operator for counting paths is exactly what we need to capture $\shl$.
\begin{theorem} \label{tqfo-shl}
	$\tqfo(\fo)$ captures $\shl$ over the class of ordered structures.
\end{theorem}
The last result perfectly illustrates the benefits of the logical framework for the descriptive complexity of counting functions. 
The distinction of the language between boolean and quantitative allows us to define operators at the later that cannot be defined at the former. 
Indeed, it is not clear how to capture $\shl$ with $\qfo(\tc)$\footnote{$\tc$ corresponds to $\fo$-logic extended with transitive closure}, or any other extension of $\qfo(\fo)$ at the boolean level.
In contrast, with $\rqfo$ one can naturally define the path operator at the quantitative level and easily capture $\shl$.
Therefore, Theorem~\ref{tqfo-shl} is good example of how to go beyond classical logic to have a better understanding of the expressivity of counting.

>>>>

In the previous sections, we use weighted logics to give a framework for descriptive complexity of counting complexity classes. Here, we go beyond weighted logics 
%to add unexplored quantifiers at the quantitative level. Specifically, we 
and give the first steps on defining recursion at the quantitative level.
%quantitative recursion. 
This goal is not trivial not only because we want to add recursion over functions, but also because 
it is not clear what could be the right notion of ``fixed point''. 
%for two reasons: (1) we want to add recursion over functions and (2) it is not clear what could be the right notion of ``fixed point'' . 
To this end, we show first how to extend $\qso$ with function symbols that are later
used 
%to later use them to define recursion at a quantitative level. 
%Furthermore, we 
to define a natural generalization for functions of 
%(boolean) 
the notion of least fixed point of LFP, which is denoted as \emph{least s-fixed point}.
% and, 
 As a proof of concept, we show that how this notion can be used to captures $\fp$.
Moreover, we use this concept
%our new tool 
to define an operator for counting paths in a graph, a natural generalization of the transitive closure operator~\cite{immerman1999descriptive}, and show that this gives rise to a logic that captures~$\shl$. 
%the expressiveness of

%For defining recursion, 
We start
%first need to define 
by defining an extension of $\qso$ with function symbols. Assume that $\fs$ is an infinite set of function symbols, where each $h \in \fs$ has an associated arity
%, which is 
denoted by $\arity(h)$. Then the set of $\fqso$ formulas over a relational signature $\R$ is defined by the following grammar:
\begin{multline}
\label{eq-fqso}
	\alpha := \varphi \ \mid \  s \  \mid \  h(x_1, \ldots, x_\ell) \  \mid \
	(\alpha \add \alpha) \  \mid\  (\alpha \mult \alpha) \  \mid \\  
	\sa{x} \alpha \  \mid \
	\pa{x} \alpha \  \mid \
	\sa{X} \alpha \  \mid \
	\pa{X} \alpha,
\end{multline}
where $h \in \fs$, $\arity(h) = \ell$ and $x_1, \ldots, x_\ell$ is a sequence of (non-necessarily distinct) first-order variables. Given an $\R$-structure $\A$ with domain $A$, we say that $F$ is a \emph{function assignment} for $\A$ if for every $h \in \fs$ with $\arity(h) = \ell$, we have that $F(h) :  A^\ell \to \N$. The notion of function assignment is used to extend the semantics of $\qso$ to the case of a quantitative formula of the form $h(x_1, \ldots, x_\ell)$. More precisely, given first-order and second-order assignments $v$ and $V$ for $\A$, respectively, 
we have that:
\begin{eqnarray*}
\sem{h(x_1, \ldots, x_\ell)}(\A,v,V,F) & = & F(h)(v(x_1),\ldots, v(x_\ell)).
\end{eqnarray*}
As for the case of $\qfo$, we define $\fqfo$ disallowing quantifiers $\Sigma X$ and $\Pi X$ in \eqref{eq-fqso}.

It is worth noting that function symbols in $\fqso$ represent functions from tuples to natural numbers, so they are different from 
%cannot be compared to 
the classical notion of function symbol in $\fo$~\cite{enderton2001mathematical}. 
%~$\shl$.first-order logics
Furthermore, %one could see 
a function symbol can be seen as an ``oracle'' that is instantiated by the function assignment. 
%As far as we know, 
To the best of our knowledge, this is the first paper to propose 
%proposing 
this extension on weighted logics, which should be further investigated. 
% and it would be interesting to study this concept further (i.e. outside recursion).

We define an 
extension of 
%operator which extends 
LFP \cite{I86,vardi1982complexity} 
%over $\fo$ to 
to allow counting. 
%We show how to define this operator in $\qfo(\fo)$ and we discuss how to extend these ideas to $\qso$ towards the end of the section. 
%Fix a relational signature $\R$. 
More precisely, the set of $\rqfo(\fo)$ formulas over a relational signature $\R$, where $\rqfo$ stands for recursive $\qfo$, is defined as an extension of $\qfo(\fo)$ that includes the formula 
%by adding the operator 
$\clfp{\beta(\x, h)}$, where (1) $\x = (x_1, \ldots, x_\ell)$ is a sequence of $\ell$ distinct first-order variables, (2) $\beta(\x, h)$ is an $\fqfo(\fo)$-formula over $\R$ whose only function symbol is $h$, and (3) $\arity(h) = \ell$. The free variables of the formula $\clfp{\beta(\x,h)}$ are $x_1, \ldots, x_\ell$; in particular, $h$ is not considered to be free in this formula.

Fix an $\R$-structure with domain $A$ and a quantitative formula $\clfp{\beta(\x,h)}$ with $\arity(h) = \ell$, and assume that $\F$ is the set of functions $f :A^\ell \to \N$. To define the semantics of $\clfp{\beta(\x,h)}$, we first show how $\beta(\x,h)$ can be interpreted as an operator $T_{\beta}$ on $\F$. More precisely, for every $f \in \F$ and tuple $\a = (a_1, \ldots, a_{\ell}) \in A^\ell$, the function $T_{\beta}(f)$ satisfies that:
\begin{eqnarray*}
T_{\beta}(f)(\a) & = & \sem{\beta(\x, h)}(\A,v,F),
\end{eqnarray*}
where $v$ is a first-order assignment  for $\A$ such that $v(x_i) = a_i$ for every $i \in \{1, \ldots, \ell\}$, and $F$ is a function assignment for $\A$ such that $F(h) = f$.

@@

\begin{example} \label{ex:count-path}
%As an example, 
We would like to define an $\rqfo(\fo)$-formula that, given a directed acyclic graph $G$ with $n$ nodes and a pair of nodes $b$, $c$ in $G$, counts the number of paths of length at most $n$ from $b$ to $c$ in $G$. To this end, assume that graphs are encoded using the relational signature $\R = \{ E(\cdot,\cdot) \}$, and then define formula $\alpha(x, y, f)$ as follows:
$$
%\alpha(x,y,R,\pi) = 
%(\neg \exists zR(z))\cdot(x = y) 
E(x,y) + \sa{z} f(x,z)\cdot E(z,y).
$$
We have that $\clfp{\alpha(x,y,f)}$ defines our counting function. In fact, assume that $\A$ is an $\R$-structure with $n$ elements in its domain encoding an acyclic directed graph. Moreover, assume that $b,c$ are elements of $\A$ and $v$ is a first-order assignment over $\A$ such that $v(x) = b$ and $v(y) = c$. Then we have that $\sem{\clfp{\alpha(x,y,f)}}(\A,v)$ is equal to the  number of paths in $\A$ from $b$ to $c$ of length at most $n$.

Assume now that we need to extend our previous counting function to the case of arbitrary directed graphs. To this end, suppose that $\varphi_{\text{\rm first}}(x)$ and $\varphi_{\text{succ}}(x,y)$ are the $\fo$-formulas for defining the first and successor predicates, respectively, of $<$. Moreover, define formula $\beta(x, y, t, g)$ as follows:
\begin{multline*}
(E(x,y) + \sa{z} g(x,z,t)\cdot E(z,y)) \cdot \varphi_{\text{\rm first}}(t) \ +\\
\sa{t'} \varphi_{\text{succ}}(t',t) \cdot \left(\sa{x'} \sa{y'} g(x',y',t') \right)
\end{multline*}
Then our extended counting function is defined by the following formula:
$$
\sa{t} (\varphi_{\text{\rm first}}(t) \wedge \clfp{\beta(x,y,t,g)}).
$$ 
In fact, the number of paths of length at most $n$ from a node $x$ to a node $y$ is recursively computed by using the formula $(E(x,y) + \sa{z} g(x,z,t)\cdot E(z,y)) \cdot \varphi_{\text{\rm first}}(t)$, which stores this value in $g(x,y,t)$ with $t$ the first element in the domain.  The other formula $\sa{t'} \varphi_{\text{succ}}(t',t) \cdot \left(\sa{x'} \sa{y'} g(x',y',t') \right)$ is just an auxiliary artifact that is used 
%The other formula, the one with the filter $\varphi_{\text{succ}}(t',t)$, is used 
as a counter to allow reaching a fixed point in the support of $g$ in $n$ steps. Notice that the use of the filter $\varphi_{\text{succ}}(t',t)$ prevents this formula for incrementing the value of $g(x,y,t)$ when $t$ is the first element in the domain.
\end{example}
%
%
%It is well known that least fixed point logic is contained in second-order logic \cite{L04}. In the following theorem we show that the same holds in our case.
%\begin{theorem} \label{so-rec}
%$\rqfo \subseteq \qso$
%%	Given a positive $\fo$ formula $\varphi(\bar{x},R)$ and a $\qfo$ formula $\alpha(\bar{x})$, there exists a $\qso$ formula $\beta(\bar{x})$ such that $\sem{[\alfp\varphi(\bar{x},R)\mid \alpha(\bar{x},R)](\bar{x})} = \sem{\beta(\bar{x})}$.
%\end{theorem} 
%Note that, 
In contrast with $\lfp$, to reach a fixed point 
%here 
we do not need to impose any positive restriction on the formula $\beta(\x,h)$.
Indeed, since $\beta$ is constructed from monotones operations (i.e. sum and product) over the natural numbers, the resulting operator $T_{\beta}$ is monotone as well.

Now that a least fixed point operator over functions is defined, the next step is to understand its expressive power.
In the following theorem, we show that this operator can be used to capture $\fp$.
%result we show how lsfp-operator can be used to capture all functions that can be computed in polynomial time.
\begin{theorem} \label{rqfo-fo-cap}
	$\rqfo(\fo)$ captures $\fp$ over 
	%the class of 
	ordered structures.
\end{theorem}

@@

Our last goal in this section is to use the new characterization of $\fp$ to explore classes below it.
It was shown in \cite{I86,I88} that $\fo$ extended with a transitive closure operator captures $\nlog$. 
%Moreover, it was shown in \cite{I83} that $\so$ extended with a transitive closure operator captures $\pspace$. 
Inspired by this work, we show that a restricted version of $\rqfo$ can be used to capture $\shl$, the counting version of $\nlog$. 
Specifically, we use $\rqfo$ to define an operator for counting the number of paths in a directed graph, which is what is needed to capture $\shl$. 
%Besides, we show that the same idea can be used to extend $\qso$ allowing to capture harder complexity classes.

Given a relation signature $\R$, the set of transitive $\qfo$ formulas ($\tqfo$-formulas) is defined as an extension of $\qfo$ with the formula
%\begin{multline}
%	\label{eq-def-tqso}
%	\alpha := \varphi \, \mid \, s \, \mid \, (\alpha \add \alpha) \, \mid\, (\alpha \mult \alpha) \, \mid \, 
%	\sa{x} \alpha \, \mid\, \
%	\pa{x} \alpha \, \mid \\ 
%	\sa{X} \alpha \, \mid \, \pa{X} \alpha \, \mid \, [\pth \psi(\bar{x}, \bar{X},\bar{y}, \bar{Y})],
%\end{multline}
$
[\pth \psi(\bar{x},\bar{y})]
$, where $\psi(\x, \y)$ is an $\fo$-formula over $\R$, and $\bar{x} = (x_1, \ldots, x_k)$, $\bar{y} = (y_1, \ldots, y_k)$ are tuples of pairwise distinct first-order variables. The semantics of $[\pth \psi(\bar{x},\bar{y})]$ can easily be defined in terms of $\rqfo(\fo)$ as follows. 
Given an $\R$-structure $\A$ with domain $A$, define a (directed) graph $\cG_{\psi}(\A) = (N,E)$ such that $N = A^k$ and for every pair $\bar b, \bar c \in N$, it holds that $(\bar b, \bar c) \in E$ if, and only if, $\A \models \psi(\bar b, \bar c)$.
Similar than for Example~\ref{ex:count-path}, we can count the paths of length at most $|A^k|$ in $ \cG_{\psi}(\A)$ with the formula $\beta_{\psi(\bar{x},\bar{y})}(\x, \y, \t, g)$:
\begin{multline*}
(\psi(\bar{x},\bar{y}) + \sa{\z} g(\x,\z,\t)\cdot \psi(\z,\y)) \cdot \varphi_{\text{\rm first-lex}}(\t) \ +\\
\sa{\t'} \varphi_{\text{succ-lex}}(\t',\t) \cdot \left(\sa{\x'} \sa{\y'} g(\x',\y',\t') \right),
\end{multline*}
where $\varphi_{\text{\rm first-lex}}$ and $\varphi_{\text{succ-lex}}$ are $\fo$-formulas defining the first and successor predicates over tuples in $A^k$, following the lexicographic order induced by~$<$.
Then the semantics of the path operator can be defined by using the following definition of $[\pth \psi(\bar{x},\bar{y})]$ in $\rqfo$:
\[
[\pth \psi(\bar{x}, \bar{y})] \ :=\ \sa{\t} (\varphi_{\text{\rm first}}(\t) \wedge \clfp{\beta_{\psi(\bar{x},\bar{y})}(\x,\y,\t,g)}).
\]
In other words, $\sem{[\pth \psi(\bar{x}, \bar{y})]}(\A,v)$ counts the number of paths from $v(\bar x)$ to $v(\bar y)$ in the graph $\cG_{\psi}(\A)$ whose length is at most~$|A^k|$.
%As for the case of $\qso$, the logic $\tqso(\LL)$ is obtained by restricting $\varphi$ in \eqref{eq-def-tqso} to be a formula in $\LL$. Moreover, the logic $\tqfo$ is obtained by disallowing in \eqref{eq-def-tqso} formulas $\sa{X} \alpha$ and $\pa{X} \alpha$, and by only allowing  first-order free-variables in the formula $\psi$ used in $[\pth \psi]$ in \eqref{eq-def-tqso}. 
As it was previously said, the operator for counting paths is exactly what we need to capture $\shl$.
\begin{theorem} \label{tqfo-shl}
	$\tqfo(\fo)$ captures $\shl$ over ordered structures.
\end{theorem}
This last result 
perfectly 
illustrates the benefits of our logical framework for the development of descriptive complexity for counting %functions.
complexity classes.  
The distinction in the language between the Boolean and the quantitative level allows us to define operators at the later level that cannot be defined at the former. 
As a example showing how fundamental this separation is, consider the issue of extending $\qfo(\fo)$ at the Boolean level in order to capture $\shl$. The natural alternative to do this is to use $\fo$ extending with a transitive closure operator, which is denoted by $\tc$. But then the problem is that for every language $L \in \nlog$, it holds that its characteristic function $\chi_L \in \qfo(\tc)$, where $\chi_L(w) = 1$ if $w \in L$, and $\chi_L(w) = 0$ otherwise. Thus, if we assume that

$\qfo(\tc)$ can def
In fact, it is not clear how $\shl$ could be capture by 
to capture $\shl$ with $\qfo(\tc)$\footnote{$\tc$ corresponds to $\fo$-logic extended with transitive closure.}, or any other extension of $\qfo(\fo)$ at the boolean level.
In contrast, with $\rqfo$ one can naturally define the path operator at the quantitative level and easily capture $\shl$.
Therefore, Theorem~\ref{tqfo-shl} is good example of how to go beyond classical logic to have a better understanding of the expressivity of counting.
